A-maze of Natural Stories: Texts are comprehensible using the Maze task

### 2022-08-16 Papers/Paper/manuscript.Rmd
Roger updated ms

To assess the shape of the RT-surprisal relationship, we fit generalized additive models (GAMs). For these models, we only included data that occurred before any mistakes in the sentence; due to limits of model vocabulary, words with punctuation and some uncommon or proper nouns were excluded. We used surprisals generated by 4 different language models for robustness. (See Methods for details on language models, exclusions, and model fit.)

@@

The main effects of current and previous word surprisals on RT are shown in Figure \@ref(fig:gam). Note that for each of the models, high-surprisal words are rare, with much of the data from words with between 0 and 15 bits of surprisal. All 4 models show a roughly linear relationship between current word surprisal and RT, especially in the region with more data.

@@

@@

# Appendix D
(ref:gam-grid) CAPTION GOES HERE

>>>>

Due to previous reports of a length--frequency interaction in RT measures [@kliegl-etal:2006], before pursuing our primary question of the functional form of the surprisal--RT relationship, as an exploratory measure we fitted generalized additive models (GAMs) with not only the main effects but also the two-way interactions between surprisal, length, and frequency, for the current word and for the previous word. This analysis revealed significant effects of current-word and previous-word surprisal, current-word and previous-word length, and significant interactions of current-word frequency by length and frequency by surprisal. The other main effects and interactions did not reach statistical significance.^[These are results from `mgcv`'s `summary()`; the $p$-values are approximate.] Appendix C provides tables and plots of these effects and interactions for GPT-2. The interactions can be summarized as long low-frequency words and surprising, high-frequency words as having especially long RTs; and surprising, low-frequency words as having shorter RTs than would otherwise be predicted. However, these effects are small in terms of variance explained compared to the current-word surprisal effect, which is by far the largest single effect in the model. For simplicity we therefore set aside the interaction terms involving surprisal for the remainder of this analysis.

To assess the shape of the RT-surprisal relationship, we then fitted generalized additive models (GAMs). For these models, we only included data that occurred before any mistakes in the sentence; due to limits of model vocabulary, words with punctuation and some uncommon or proper nouns were excluded. We used surprisals generated by 4 different language models for robustness. (See Methods for details on language models, exclusions, and model fit.)

@@

The main effects of current and previous word surprisals on RT are shown in Figure \@ref(fig:gam). Note that for each of the models, high-surprisal words are rare, with much of the data from words with between 0 and 15 bits of surprisal. All 4 models show a roughly linear relationship between current word surprisal and RT, especially in the region with more data. To determine the goodness of fit of a model in which word probability effects on RT are taken to be linear in surprisal, we also fitted GAM models with both parametric linear and nonparametric non-linear terms for surprisal; for all but the n-gram model, these analyses supported a linear effect of surprisal (Appendix D).

@@

We use `mgcv`'s `ti()` tensor interaction terms to test all main effects and two-way interactions among frequency, length, and surprisal for the current word and for the previous word. These effects are visualized in Figure \@ref(fig:gam-grid) and `mgcv`'s approximate significance levels are give in Table XXX insert table XXX. Based on these approximate significance levels, the main effects of current and previous word surprisal and length are significant, as are the current-word frequency-by-length and frequency-by-surprisal interactions; other terms are not statistically significant. These significant interactions can be summarized as especially long, infrequent words being especially slow to select; especially frequent and surprising words being especially slow to select; and especially infrequent and surprising words being less slow to select than a main-effects-only model would predict. The data driving these interactions are in the sparse tails of the word length and surprisal distributions, and as the $F$ statistics in Table XXX show, their variance explained is small relative to the large effect of current-word surprisal, so in the main-text analysis we set these interactions aside.

XXX Veronica could we show the "approximate significance of smooth terms" results for gam_gpt_3 in a table in this Appendix as well?

(ref:gam-grid) Generalized Additive Model main effects and two-way interactions among frequency, length, and surprisal for A-Maze reading of the Natural Stories corpus. Confidence bands do not take into account the uncertainty associated with `mgcv` hyperparameter estimation (@wood:2006GAMs).

# Appendix D

The `mgcv` package's implementation of Generalized Additive Models [@wood:2006GAMs] allows linear and nonparametric spline effects of the same continuous predictor to be entered simultaneously into a model. Doing so associates only the nonlinear part of the effect to the spline term, allowing for approximate statistical testing of the linear and non-linear components of the effect respectively. We thus test for whether the effect of surprisal on A-Maze RTs is best described as linear or includes a non-linear component, using the `mgcv` formula:

`rt ~ surprisal + s(surprisal, bs="cr", k=20) + ti(freq, bs="cr") + ti(len, bs="cr") + prev_surp + s(prev_surp, bs="cr", k=20) + ti(prev_freq, bs="cr") + ti(prev_len, bs="cr")`

The results are in Table \@ref(tab:gam-linearity-table). For all but the 5-gram surprisal estimate, there is overwhelming evidence for a linear contribution of current-word surprisal, but little to no evidence for a non-linear contribution. For the 5-gram estimate, there is overwhelming evidence for the linear term, and some evidence for a nonlinearity as well. Consulting \@ref(fig:gam) shows that this nonlinearity takes the form of the surprisal effect dwindling to zero in the sparse tail of high-surprisal words. This nonlinearity is plausibly due to the measurement error (high variance) of using counts to estimate very low multinomial probabilities. Taken together with the 5-gram model's inferior overall fit, we conclude from this analysis that the evidence is quite strong that, like self-paced reading and eye tracking, A-Maze reading of naturalistic texts exhibits a linear effect of surprisal on RTs.

@@
